---
title: "External validation using `pmcalibration`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{External validation using `pmcalibration`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pmcalibration)

# simulate some data for vignette
set.seed(2345)
dat <- sim_dat(1000, a1 = -3, a3 = .3)

# show the first 3 columns (col 4 is the true linear predictor/LP)
head(dat[-4])
```

We have data with a binary outcome, `y`, and two 'predictor' variables, `x1` and `x2`. Suppose we have an existing model for predicting `y` from `x1` and `x2` that is as follows

```
p(y = 1) = plogis( -3 + 1*x1 + 1*x2 )
```

To externally validate this model on this new data we need to calculate the predicted probabilities. We'll also extract the observed outcomes.

```{r}
p <- plogis(with(dat, -3 + x1 + x2))
y <- dat$y
```

First we can check 'calibration-in-the-large' via the calibration intercept and slope. 

```{r}
logistic_cal(y = y, p = p)
```

The calibration-intercept suggests no particular bias with a point estimate not far off zero. The calibration slope suggests that predicted probabilities are too extreme. However, this logistic calibration enforces a linear relationship between logit transformed probabilities and the log odds of y = 1. 

Below we use `pmcalibration` to fit a flexible calibration curve, allowing for a non-linear relationship between predicted and actual probabilities. This assesses 'moderate calibration' according to the heirarchy of [Van Calster et al. (2016)](https://pubmed.ncbi.nlm.nih.gov/26772608/).

In the example below, we fit a calibration curve using `mgcv::gam` via a penalized thin plate regression spline (see `?mgcv::tprs`). `pmcalibration` calculates various metrics from the absolute difference between the predicted probability and the actual probability (as estimated by the calibration curve). In this case 95% confidence intervals for these metrics are calculated via simulation based inference. 

```{r}
(cc <- pmcalibration(y = y, p = p, 
                     smooth = "gam", bs = "tp", 
                     k = 10, logitp = T, 
                     ci = "sim", method="REML"))

```
 
The printed metrics can be interpreted as follows: 

- `Eavg` suggests that the average difference between prediction and actual probability of the outcome is 0.02 (or 2%) with a 95% CI of [0.009, 0.031].
- `E50` is the median difference between prediction and observed probability (inferred from calibration curve). 50% of differences are 0.013 or smaller.
- `E90` is the 90th percentile difference. 90% of differences are 0.024 or smaller. 
- `Emax` is the largest observed difference between predicted and observed probability. The model can be off by up to 0.21, with a broad confidence interval.
- `ECI` is the average squared difference between predicted and observed probabilities (multiplied by 100). See [Van Hoorde et al. (2015)](https://pubmed.ncbi.nlm.nih.gov/25579635/).

A quick and simple plot of the calibration curve, and 95% confidence interval, can be obtained via `plot`.

```{r, fig.height=5, fig.width=6}
plot(cc)
```

Or one could use `get_cc` to extract data for plotting with method of your choice. The plot below also shows the distribution of predicted probabilities.

```{r, fig.height=5, fig.width=6}
library(ggplot2)

get_cc(cc) |>
  ggplot(aes(x = p, y = p_c, ymin=lower, ymax=upper)) +
  geom_abline(intercept = 0, slope = 1, lty=2) +
  geom_line() +
  geom_ribbon(alpha = 1/2, fill="lightblue") +
  coord_cartesian(xlim=c(0,1), ylim=c(0,1)) +
  labs(x = "Predicted", y = "Estimated") +
  theme_bw(base_size = 14) +
  geom_histogram(data = data.frame(p = p), aes(x=p, y=after_stat(density)*.01),
                 binwidth = .001, inherit.aes = F, alpha=1/2)

```

The model in its current form very slightly underestimates risk at low levels of predicted risk and then overestimates risk at predicted probabilities of over 0.4.

The results above can be compared with `rms::val.prob`. Note that this uses `lowess(p, y, iter=0)` to fit a non-linear (nonparametric) calibration curve. This calibration curve suggests that the overestimation at high levels of predicted risk is even more extreme that that suggested by `gam` calibration curve above. This is particularly evident in the estimate of `Emax` (0.35 vs 0.21). 

```{r, fig.height=5, fig.width=6}
rms::val.prob(p = p, y = y)
```

Note also that the calibration intercept reported by `rms::val.prob` comes from the same logistic regression as that used to estimate the calibration slope. In `logistic_cal` the calbration intercept is estimated via a `glm` with logit transformed predicted probabilities included as an offset term (i.e., with slope fixed to 1 - see, e.g., [Van Calster et al., 2016](https://pubmed.ncbi.nlm.nih.gov/26772608/)). The calibration slope is estimated via a separate `glm`.

\
\
\
